---
title: "GDP per capita and life expectancy in 2020"
output:
  html_document:
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  dpi = 120
)
library(ggplot2)
library(maps)
library(scales)
```

This report uses Gapminder-style country panels for **GDP per capita** (`gdp_pcap.csv`) and **life expectancy at birth** (`lex.csv`), restricted to **2020**.

## Data structure

```{r peek}
gdp_raw <- read.csv("../gdp_pcap.csv", check.names = FALSE, stringsAsFactors = FALSE)
lex_raw <- read.csv("../lex.csv", check.names = FALSE, stringsAsFactors = FALSE)
yr <- "2020"
gdp_sample <- gdp_raw[seq_len(min(5, nrow(gdp_raw))), c("geo", "name", "1800", "1950", "2000", yr)]
lex_sample <- lex_raw[seq_len(min(5, nrow(lex_raw))), c("geo", "name", "1800", "1950", "2000", yr)]
```

Each file is in **wide** form: `geo` (3-letter code), `name`, and one column per year. The tables below show the first five countries and a few year columns.

```{r show-samples, results="asis"}
cat("**Sample: GDP per capita** (first rows, selected years)\n\n")
knitr::kable(gdp_sample)
cat("\n\n**Sample: Life expectancy** (first rows, selected years)\n\n")
knitr::kable(lex_sample)
```

## 2020 subset

```{r subset-2020}
gdp_2020 <- data.frame(
  geo = toupper(gdp_raw[["geo"]]),
  name = gdp_raw[["name"]],
  gdp_pcap = suppressWarnings(as.numeric(gdp_raw[[yr]])),
  stringsAsFactors = FALSE
)
lex_2020 <- data.frame(
  geo = toupper(lex_raw[["geo"]]),
  life_exp = suppressWarnings(as.numeric(lex_raw[[yr]])),
  stringsAsFactors = FALSE
)

combined_2020 <- merge(gdp_2020, lex_2020, by = "geo", all = FALSE)
combined_2020 <- combined_2020[
  !is.na(combined_2020$gdp_pcap) & !is.na(combined_2020$life_exp),
]
n_countries <- nrow(combined_2020)
```

After dropping rows with missing 2020 values for either indicator, the analysis uses `r n_countries` countries with both GDP per capita and life expectancy in 2020.

## World maps (2020)

Country values are matched to map polygons via `maps::iso3166` (ISO alpha-3 to map region name) and `ggplot2::map_data("world")`. Unmatched or missing countries appear in light grey.

```{r world-maps}
iso <- data.frame(
  geo = iso3166[["a3"]],
  map_region = iso3166[["mapname"]],
  stringsAsFactors = FALSE
)

gdp_map <- merge(iso, gdp_2020[, c("geo", "gdp_pcap")], by = "geo", all.x = FALSE)
lex_map <- merge(iso, lex_2020, by = "geo", all.x = FALSE)

world <- map_data("world")

map_gdp_df <- merge(world, gdp_map, by.x = "region", by.y = "map_region", all.x = TRUE)
map_lex_df <- merge(world, lex_map, by.x = "region", by.y = "map_region", all.x = TRUE)

map_gdp_plot <- ggplot(map_gdp_df, aes(x = long, y = lat, group = group, fill = gdp_pcap)) +
  geom_polygon(color = NA) +
  coord_fixed(1.3) +
  scale_y_continuous(limits = c(-60, 90), expand = c(0, 0)) +
  scale_fill_viridis_c(
    option = "plasma",
    trans = "log10",
    na.value = "grey90",
    labels = label_number(scale_cut = cut_long_scale()),
    name = "GDP per capita\n(constant intl. $, log scale)"
  ) +
  theme_void(base_size = 12) +
  theme(legend.position = "right") +
  labs(title = "GDP per capita, 2020", subtitle = "World map (Natural Earth via maps); missing = grey")

map_lex_plot <- ggplot(map_lex_df, aes(x = long, y = lat, group = group, fill = life_exp)) +
  geom_polygon(color = NA) +
  coord_fixed(1.3) +
  scale_y_continuous(limits = c(-60, 90), expand = c(0, 0)) +
  scale_fill_viridis_c(
    option = "cividis",
    na.value = "grey90",
    name = "Life expectancy\n(years at birth)"
  ) +
  theme_void(base_size = 12) +
  theme(legend.position = "right") +
  labs(title = "Life expectancy at birth, 2020", subtitle = "World map (Natural Earth via maps); missing = grey")

map_gdp_plot
map_lex_plot
```

## Relationship in 2020

```{r scatter}
cor_gdp_lex <- cor(combined_2020$gdp_pcap, combined_2020$life_exp, method = "pearson")

scatter_plot <- ggplot(combined_2020, aes(x = gdp_pcap, y = life_exp)) +
  geom_point(alpha = 0.65, size = 2, color = "#2C7FB8") +
  geom_smooth(method = "lm", se = TRUE, color = "#253494", linewidth = 0.6) +
  scale_x_log10(labels = label_number(scale_cut = cut_long_scale())) +
  labs(
    title = "GDP per capita vs life expectancy, 2020",
    x = "GDP per capita (log10 scale, constant international $)",
    y = "Life expectancy at birth (years)"
  ) +
  theme_minimal(base_size = 12)

scatter_plot
```

Pearson correlation (on raw GDP values): **`r round(cor_gdp_lex, 3)`**.

```{r summary-table, results="asis"}
sum_tbl <- data.frame(
  n = nrow(combined_2020),
  median_gdp = median(combined_2020$gdp_pcap),
  median_lex = median(combined_2020$life_exp),
  min_gdp = min(combined_2020$gdp_pcap),
  max_gdp = max(combined_2020$gdp_pcap),
  min_lex = min(combined_2020$life_exp),
  max_lex = max(combined_2020$life_exp)
)
cat("**Summary of paired 2020 observations**\n\n")
knitr::kable(sum_tbl, digits = 1)
```

## Interpretation

In 2020, **higher GDP per capita is associated with higher life expectancy** across countries: the cloud of points slopes upward, especially on a log scale for income, and the fitted line on the log scale is clearly positive. Richer countries cluster toward the top-right; many low-income countries have lower life expectancy, though there is **meaningful dispersion** (public health, inequality, institutions, and recent shocks all move countries off a simple income–longevity line).

## Reflection 
Interacting with cursor was easy. I entered the prompts I needed to create the report with the dataset. The result is amazing and so much time was saved using this tool.